A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-11-11 jour/jour.tex
Paragraph on model construction.

For \rn{Unique}, Figure~\ref{fig:uniqueness-algo} considers a partition $\mathcal{U}$ of the equivalence classes of our set $\Ec$
such that codatatype equivalence classes having a constructor term with top symbol $\const{C}$ are placed in the same subset, 
for each constructor $\const{C}$.
We write $\ec{t} =_{\mathcal{U}} \ec{u}$ to denote that $\ec{t}$ and $\ec{u}$ reside in the same subset within $\mathcal{U}$.
We then refine $\mathcal{U}$ by 
constructing a partition $\mathcal{U}_i$ of each $U_i \in \mathcal{U}$ such that distinct $\ec{t}$ and $\ec{u}$ reside in the same subset of $\mathcal{U}_i$ 
if and only if they contain constructor terms $\const{C}( t_1, \ldots, t_k )$ and $\const{C}( u_1, \ldots, u_k )$
where each $t_j, u_j$ are such that $\ec{t_j}$ and $\ec{u_j}$ reside in the same subset of $\mathcal{U}$ for each $j=1, \ldots, k$.
Let $\mathcal{U'}$ be the union of each of these partitions.
We update $\mathcal{U}$ to $\mathcal{U'}$.
This refinement is repeated until $\mathcal{U}$ is left unchanged.
If the resulting partition $\mathcal{U}$ contains a subset $U$ having two distinct equivalence classes $\ec{t}$ and $\ec{u}$, then \rn{Unique} applies to $t$ and $u$.
\textbf{Mention this coincides with DFA minimization algorithm.}

@@

The implementation is able to produce models for satisfiable inputs.
As described in Section~\ref{sec:termination-and-correctness},
given a saturated set $\Ec$, we may construct a map $\ValC$ from variables (corresponding to the equivalence classes of $\Ec$) to closed normal $\MU$-terms.
%In practice, this requires a fair enumerator for each (co)datatype $\tau$.

%Recall that to assign a value $\MU {\Varec{x}}.\; t$ to variable

%We do not explicitly construct the set $\BAD_{\smash{\J}}^x( \Val )$.
%Instead,

>>>>

For \rn{Unique}, Figure~\ref{fig:uniqueness-algo} considers a partition $\mathcal{U}$ of the equivalence classes of our set $\Ec$
such that codatatype equivalence classes having a constructor term with top symbol $\const{C}$ are placed in the same subset, 
for each constructor $\const{C}$.
We write $\ec{t} =_{\mathcal{U}} \ec{u}$ to denote that $\ec{t}$ and $\ec{u}$ reside in the same subset within $\mathcal{U}$.
We then refine $\mathcal{U}$ by 
constructing a partition $\mathcal{U}_i$ of each $U_i \in \mathcal{U}$ such that distinct $\ec{t}$ and $\ec{u}$ reside in the same subset of $\mathcal{U}_i$ 
if and only if they contain constructor terms $\const{C}( t_1, \ldots, t_k )$ and $\const{C}( u_1, \ldots, u_k )$
where each $t_j, u_j$ are such that $\ec{t_j}$ and $\ec{u_j}$ reside in the same subset of $\mathcal{U}$ for each $j=1, \ldots, k$.
Let $\mathcal{U'}$ be the union of each of these partitions.
We update $\mathcal{U}$ to $\mathcal{U'}$.
This refinement is repeated until $\mathcal{U}$ is left unchanged.
If the resulting partition $\mathcal{U}$ contains a subset $U$ having two distinct equivalence classes $\ec{t}$ and $\ec{u}$, then \rn{Unique} applies to $t$ and $u$.
This algorithm is similar to the classic algorithm for minimizing deterministic finite automata (DFA). \textbf{maybe cite?}

@@

When instructed, the implementation produces models for satisfiable inputs.
As described in Section~\ref{sec:termination-and-correctness},
given a saturated set $\Ec$, we may construct a map $\ValC$ from 
the equivalence classes of $\Ec$ to closed normal $\MU$-terms.
Recall our construction of $\ValC$ requires picking values $\MU {\Varec{x}}.\; t$ for each $\Varec{x} \in \FV( \ValC )$
such that $\MU {\Varec{x}}.\; t$ does not occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
To pick values $\MU {\Varec{x}}.\; t$, we use a fair enumerator for each (co)datatype $\tau$ which enumerates the values
of type $\tau$ in a normal interpretation, which are closed normal $\MU$-terms.
Concretely, we enumerate a stream of \emph{every} $\MU$-term of our signature $\Sigma$ in a fair manner,
discarding whose that have free variables or are not normal.
For those that are closed and normal, we then check whether they occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$,
which we do not explicitly construct.
%We do not explicitly construct the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
Instead, to check if a term $\MU {\Varec{x}}.\; t$ occurs in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$, 
we check if it is alpha-equivalent to a closed-term $(\MU {\Varec{y}}.\; u) \{ \Varec{x} \mapsto z \}$ for
some $\MU {\Varec{y}}.\; u$ in the range of $\ValC$ and variable $z$.
This can be efficiently checked by matching the term $\MU {\Varec{x}}.\; t$ with the term $\MU {\Varec{y}}.\; u$.

%Once the mapping $\ValC$ is constructed, our model $\J$ interprets
%$\J( x ) = \ValC \ec{x}$ for each variable $x$ occurring free in $\Ec$.
